### 2022-10-12 AISTATS2023/paper.tex
latest

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

@@

% Such ambiguities could be resolved if the heterogeneity of instance-level feature effects is provided in addition to the average effect. 
% In this paper, we measure the heterogeneity as the standard deviation of the instance-level effects and we refer to it using as the \emph{uncertainty of the effect}.

%Therefore, global methods should quantify not only the aggregated (averaged) effect, but also the level of heterogeneity of instance-level explanations.

%There are three widely-used FE methods: \emph{Partial Dependence Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
%(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
%(ALE)\citep{apley2020visualizing}. ALE has been established as the
%state-of-the-art for measuring the average effect, since PDP and MP
%have been criticized~\citep{Gromping2020MAEP} to compute erroneous
%effects when the input features are (highly) correlated.

@@

% %However, ALE faces two limitations, the first concerns ALE definition and the second ALE approximation. Regarding the definition, ALE does not inform the user about the uncertainty. 
% \color{blue}\todo{This should go to related work}In contrast, in PDP the
% heterogeneity of instance-level effects can be identified by exploring
% the Individual Conditional Expectations
% (ICE)~\citep{goldstein2015peeking}. \color{black}
% %Regarding the approximation, ALE requires partitioning the feature domain into non-overlapping intervals (bin-splitting) and estimate the effect inside each interval (bin-effect) from the training set instances that fall inside. 
% \color{blue}\todo{to be moved in related work}
% Finding
% an appropriate sequence of bins is of particular importance, since
% ALE's interpretation is meaningful only inside each
% bin~\citep{molnar2022}. However, this crucial step has not raised the
% appropriate attention. With the current
% approaches~\citep{apley2020visualizing, gkolemis22}, the domain is
% split \emph{blindly}, i.e.~without considering the characteristics of
% the instance-level effects, in \(K\) equisized bins.\color{black}

%In this paper, we present Uncertainty-aware Accumulated Local Effects(UALE), a method that visualizes in a single plot both the av erage feature effect and the uncertainty, i.e.~the level of heterogeneity of instance-level effects. The average effect is defined as in ALE and the uncertainty inherits ALE's advantages, i.e.~takes into account the correlations between features. We also propose an accurate UALE approximation. We formally prove that for an unbiased estimation of the uncertainty, the partitioning of the feature domain must respect specific conditions. Using that, we set-up an optimization problem where we search for a partitioning, a sequence of \emph{variable-size}intervals, (a) with enough instances (accurate estimation) and (b) that respect the above conditions (unbiased estimation). Finally, we propose a method for automatically finding the optimal partitioning. We apply our method in multiple synthetic and real world datasets for evaluating its performance. 
The contributions of this
paper are:
\begin{itemize}
\item A global feature effect method (UALE) that quantifies both the
  average effect and its uncertainty, i.e.,~the deviation of
  instance-level effects.
\item An unbiased approximation of the uncertainty through a dynamic
  variable-size bin partitioning of the feature domain.
%A formal proof of the conditions the feature domain partitioning must fulfill for an unbiased approximation of the uncertainty.
\item The evaluation of UALE on both synthetic and real datasets.
%An accurate UALE approximation by automatically partitioning into variable-size intervals that respect the above conditions.
\end{itemize}

@@

% as
% input and perform an equally-sized bin splitting of the attribute
% domain. The choice of \(K\) is therefore critical and introduces
% significant limitations. Specifically, setting \(K\) to a small value
% may hide fine-grain effects due to large bins while setting \(K\) to a
% high value leads to poor bin-effect estimations due to few samples per
% bin. In general, the user may receive contradictory explanations for
% different \(K\) values without a clue for which one to trust.

@@

In Section~\ref{sec:UALE-definition-1} we define UALE and we
provide an interval-based formulation that partitions the feature
domain into non-overlapping intervals (bin-splitting) and defines the
bin-effect and the bin-uncertainty. Then, we present UALE
approximation (Section~\ref{sec:UALE-approximation}), formally proving
the bin-splitting requirements for an unbiased estimation of the
uncertainty. Based on this, we present an optimization method to
determine an appropriate variable-width partitioning
(Section~\ref{sec:bin-spliting}).

@@

For the interval-based formulation, we define the bin-effect
\(\mu(z_1, z_2)\) and the corresponding bin-uncertainty
\(\sigma(z_1, z_2)\) as:

@@

%
Intuitively, if we randomly draw a point \(z^*\) given a uniform
distribution \(z^* \sim \mathcal{U}(z_1, z_2)\), the bin-effect
(Eq.~(\ref{eq:mu_bin})) and the bin-uncertainty
(Eq.~(\ref{eq:var_bin})) are the expected average effect and the
expected uncertainty, respectively. Also, if we define \(\mathcal{Z}\)
to be the sequence of \(K+1\) points that partition the domain of the
\(s\)-th feature into \(K\) variable-size intervals, i.e.
\(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can reformulate UALE to:
\begin{equation}
  \label{eq:ALE_2}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%
where \(k_x\) is the index of the bin such that
\( z_{k_x - 1} \leq x_s < z_{k_x}\).

@@

%
\begin{equation}
  \label{eq:var_bin_approx}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_1, z_2) \right )^2
\end{equation}
%
In the Appendix~\ref{sec:proof-1}, we show that
\(\hat{\sigma}^2(z_{k-1}, z_k)\) is an unbiased estimator of
\(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{\xc|x_s=z}
  \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
  z_1} \)
%
which, in the general case, overestimates
\(\sigma^2(z_1, z_2)\), as shown in Theorem~\ref{sec:theorem-1}.

\begin{theorem}
  \label{sec:theorem-1}
  If we define (a) the residual \(\rho(z)\) as the difference between
  the expected effect at \(x_s\) and the regional effect, i.e
  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), and (b) \(\mathcal{E}(z_1, z_2)\) as the
  mean squared residual of the bin, i.e.
  \(\mathcal{E}(z_1, z_2) = \frac{\int_{z_1}^{z_2}\rho^2(z) \partial
    z}{z_2 - z_1}\), then it holds
\begin{equation}
    \label{eq:bin-uncertainty-proof}
 \sigma_*^2(z_1, z_2) = \sigma^2(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
\end{equation}
  \end{theorem}

@@

Based on Theorem\ref{sec:theorem-1}, the estimation is unbiased only
when \(\mathcal{E}^2(z_1, z_2) = 0\).

@@

The quality of interval-based approximation is affected by (a) the
population of samples in each bin and (b) the error term
\(\mathcal{E}(z_1, z_2)\) of each bin. On the one hand, we favor a
partitioning of the feature domain in large-bins to allow a robust
estimation of \(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\) from a
sufficient population of samples. On the other hand, we should
minimize the aggregation of bin errors, i.e.,
\( \mathcal{E}^2_{\mathcal{Z}} = \sum_{k=1}^K\mathcal{E}^2(z_1, z_2)
\Delta z_k\), where \(\mathcal{Z} = \{z_0, \cdots, z_K\}\) and
\(\Delta z_k = z_k - z_{k-1}\). For convenience, we refer to
\(\mathcal{E}^2_{\mathcal{Z}}\) as aggregated error. We search for a
partitioning \(\mathcal{Z}_*\) that balances this trade-off.

@@

\begin{corollary}
  \label{sec:coroll}
  The minimization of the aggregation of bin variances is equivalent
  to the minimization of the aggregated error:
  \begin{equation}
  \begin{gathered}
    \label{eq:2}
    \argmin_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \mathcal{E}^2_{\mathcal{Z}}  = \argmin_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k 
  \end{gathered}
  \end{equation}
\end{corollary}

@@

\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \tau_k \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - \alpha \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
The objective \(\mathcal{L}\) searches for the partitioning
\(\mathcal{Z}_* = \{ z_0, \ldots, z_K \} \) that minimizes the
aggregated error \(\mathcal{E}_{\mathcal{Z}}^2\), while the term
\(\tau_K\) favors the partitioning that contains bigger bins, in case
many partitionings have similar aggregated error. The constraint of at
least \(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
\textit{robust} estimation. The user can choose to what extent they
favor the creation of wide bins through (a) the parameter \(\alpha\)
that controls the discount \(\tau_k\) and (b) the parameter
\(N_{\mathtt{PPB}}\) that sets the minimum population per bin. For
providing a rough idea, in our experiments we set \(\alpha = 0.2\)
which means that the discount ranges between \([0\%, 20\%]\) and
\(N_{\mathtt{PPB}} = \frac{N}{20}\).

@@

>>>>

@@

@@

The contributions of this
paper are:
\begin{itemize}
\item A global feature effect method (UALE) that quantifies both the
  average effect and its uncertainty, i.e.,~the deviation of
  instance-level effects.
\item An unbiased approximation of the uncertainty through a dynamic
  variable-size bin partitioning of the feature domain.
\item The evaluation of UALE on both synthetic and real datasets.
\end{itemize}

@@

@@

We define UALE and provide an interval-based formulation that
partitions the feature domain into non-overlapping intervals
(bin-splitting) and defines the bin-effect and the
bin-uncertainty. Then, we present UALE approximation, formally proving
the bin-splitting requirements for an unbiased estimation of the
uncertainty. Based on this, we present an optimization method to
determine an appropriate variable-width partitioning.

@@

For the interval-based formulation, we define the bin-effect
\(\mu(z_1, z_2)\) and the bin-uncertainty \(\sigma(z_1, z_2)\) as:

@@

%
Intuitively, if we randomly draw a point \(z^*\) from a uniform
distribution \(z^* \sim \mathcal{U}(z_1, z_2)\), the bin-effect
(Eq.~(\ref{eq:mu_bin})) and the bin-uncertainty
(Eq.~(\ref{eq:var_bin})) are the expected average effect and the
expected uncertainty, respectively. Also, if we denote as
\(\mathcal{Z}\) the sequence of \(K+1\) points that partition the
domain of the \(s\)-th feature into \(K\) variable-size intervals,
i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can reformulate UALE
to:
\begin{equation}
  \label{eq:ALE_2}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%
where \(k_x\) is the index of the bin such that
\( z_{k_x - 1} \leq x_s < z_{k_x}\). Eq.~(\ref{eq:ALE_2}) and
Eq.~(\ref{eq:ALE_accumulated_var}) are the interval-based formulation
of UALE.

@@

%
\begin{equation}
  \label{eq:var_bin_approx}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_1, z_2) \right )^2
\end{equation}
%
In the Appendix~\ref{sec:proof-1}, we show that
\(\hat{\sigma}^2(z_1, z_2)\) is an unbiased estimator of
\(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{\xc|x_s=z}
  \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
  z_1} \). In Theorem~\ref{sec:theorem-1} we prove that in the general
case, \(\sigma_*^2(z_1, z_2) \geq \sigma^2(z_1, z_2)\), so
\(\hat{\sigma}^2(z_1, z_2)\) is an overestimation of the bin
uncertainty.

\begin{theorem}
  \label{sec:theorem-1}
  If we define (a) the residual \(\rho(z)\) as the difference between
  the expected effect at \(z\) and the bin-effect, i.e
  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), and (b)
  \(\mathcal{E}(z_1, z_2)\) as the mean squared residual of the bin,
  i.e.
  \(\mathcal{E}(z_1, z_2) = \frac{\int_{z_1}^{z_2}\rho^2(z) \partial
    z}{z_2 - z_1}\), then it holds
\begin{equation}
    \label{eq:bin-uncertainty-proof}
 \sigma_*^2(z_1, z_2) = \sigma^2(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
\end{equation}
  \end{theorem}

@@

We refer to \(\mathcal{E}^2(z_1, z_2)\) as bin-error. Based on
Theorem\ref{sec:theorem-1}, the estimation is unbiased only when
\(\mathcal{E}^2(z_1, z_2) = 0\).

@@

The quality of UALE approximation is affected by (a) the population of
samples in each bin and (b) the error term \(\mathcal{E}(z_1, z_2)\)
of each bin. On the one hand, we favor a partitioning of the feature
domain in large-bins to allow a robust estimation of
\(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\) from a sufficient
population of samples. On the other hand, we want to minimize the
cummulative bin-error, i.e.,
\( \mathcal{E}^2_{\mathcal{Z}} = \sum_{k=1}^K\mathcal{E}^2(z_1, z_2)
\Delta z_k\), where \(\mathcal{Z} = \{z_0, \cdots, z_K\}\) and
\(\Delta z_k = z_k - z_{k-1}\). For convenience, we refer to
\(\mathcal{E}^2_{\mathcal{Z}}\) as accumulated error. We search for a
partitioning \(\mathcal{Z}_*\) that balances this trade-off.

@@

\begin{corollary}
  \label{sec:coroll}
  A bin-splitting \(\mathcal{Z}\) minimizes the accumulated if and
  only if it also minimizes the accumulated error.
  \begin{equation}
  \begin{gathered}
    \label{eq:2}
    \argmin_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \mathcal{E}^2_{\mathcal{Z}}  = \argmin_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k 
  \end{gathered}
  \end{equation}
\end{corollary}

@@

\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \tau_k \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - \alpha \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
The objective \(\mathcal{L}\) searches for a partitioning
\(\mathcal{Z}_* = \{ z_0, \ldots, z_K \} \) with low aggregated error
\(\mathcal{E}_{\mathcal{Z}}^2\). In case of many partitionings with
similar aggregate-error, the term \(\tau_K\) favors the partitioning
with more points per bin (with wider bins). The constraint of at least
\(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
\textit{robust} estimation. The user can choose to what extent they
favor the creation of wide bins through (a) the parameter \(\alpha\)
that controls the discount \(\tau_k\) and (b) the parameter
\(N_{\mathtt{PPB}}\) that sets the minimum population per bin. For
providing a rough idea, in our experiments we set \(\alpha = 0.2\)
which means that the discount ranges between \([0\%, 20\%]\) and
\(N_{\mathtt{PPB}} = \frac{N}{20}\), where \(N\) is the dataset size.

For solving the optimization problem of~Eq.\ref{eq:opt} we add some
constraints. First, we set a threshold \(K_{max}\) on the maximum
number of bins which, in turn, defines the minimum bin-width,
i.e. \(\Delta x_{min} = \frac{x_{s, max} - x_{s,min}}{K_{max}}\).
Based on that, we restrict the bin-limits to be multiples of the
minimum width, i.e.
\(z_k = k\cdot \Delta x_{min}, \text{where } k \in \{0 , \cdots,
K_{max}\} \). In this discretized solution space, we find the global
optimal using dynamic programming.

@@

\section{Dynamic Programming}
\label{sec:dynamic-programming}
